---
title: "A Critical Review of Shaher & Sayers: Posture and Enlarged Protuberances"
output: github_document
---
<h2 style="text-align:center;">Maya Arnott</h2>
<h3 style="text-align:center;">October 22, 2025</h3>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(knitr)
library(patchwork)
```



Shahar and Sayer's paper examined 1,200 lateral cervical radiographs of people aged 18 to 86. They measured the size of the external occipital protuberance (EOP) to identify if it was enlarged (defined as the EEOP >10 mm) and examined how well factors like sex, age and forward head protraction (FHP) predicted EEOP. The most surprising predictor found in this study was age; they found an association between increased age and a decrease in enthesophyte size. Here, we will investigate the validity of their claims by reviewing the data.

```{r}
# importing the dataset and doing initial tidying
exo_df = (
  read_excel("docs/data/p8105_mtp_data.xlsx", sheet = "this one", skip = 8) |>  
    janitor::clean_names() |> 
    rename(
      eop_size_category = eop_size
    ) |> 
    mutate( 
      eop_size_mm = replace_na (eop_size_mm, 0)
          )
)
```


```{r}
age_map <- c(
  "1" = "0-17",
  "2" = "18-30",
  "3" = "31-40",
  "4" = "41-50",
  "5" = "51-60",
  "6" = "60+",
  "7" = "60+",
  "8" = "60+"
)

sex_map <- c("1" = "Male", "0" = "Female")

exo_df <- exo_df |> 
  mutate(
    # convert sex and age group
    sex = factor(recode(as.character(sex), !!!sex_map)),
    age_group = factor(recode(as.character(age_group), !!!age_map),
                       levels = c("18-30","31-40","41-50","51-60","60+"),
                       ordered = TRUE),
    
    # numeric transformations
    age = as.numeric(age),
    eop_size_mm = as.numeric(eop_size_mm),
    fhp_size_mm = as.numeric(fhp_size_mm),
    
    # factor transformations
    eop_size_category = factor(eop_size_category, levels = 0:5, ordered = TRUE),
    eop_visibility_classification = as.factor(eop_visibility_classification),
    eop_shape = as.factor(eop_shape),
    fhp_category = factor(fhp_category, levels = 0:max(fhp_category, na.rm = TRUE), ordered = TRUE)
  ) |> 
  # filter out 0-17 age group
  filter(!is.na(age_group))

```


```{r}
sum_table <- exo_df |> 
  # calculates overall summary stats
  summarize(
    total_count = n(), 
    total_mean_age = mean(age, na.rm = TRUE),
    total_sd_age = sd(age, na.rm = TRUE),
    total_min_age = min(age, na.rm = TRUE),
    total_max_age = max(age, na.rm = TRUE)
  ) |> 
  # add a total row to aggregate the values
  mutate( sex = "Total") |> 
  select(sex, 
         count = total_count, 
         mean_age = total_mean_age, 
         sd_age = total_sd_age, 
         min_age = total_min_age, 
         max_age = total_max_age) |> 
  
  bind_rows(
    exo_df |> 
      # groups the stats by sex
      group_by(sex) |> 
      summarize(
        count = n(), 
        mean_age = mean(age, na.rm = TRUE),
        sd_age = sd(age, na.rm = TRUE),
        min_age = min(age, na.rm = TRUE),
        max_age = max(age, na.rm = TRUE),
        .groups = "drop"
      )
  ) |> 
  # compute % of participants by sex
  mutate(
    percent = round(count / sum(count) * 100, 1)) |> 
  select(
    sex, count, percent, mean_age, min_age, max_age)
```

First, we tidy the `exo_df` data by standardizing the names of columns. Then, we replace the missing values in column `eop_size_mm` with 0, as indicated in the header. `sex` was converted to a factor, and `age_group`, `eop_size_category`, and `fhp_category` were all converted to ordered factors to preserve the natural ranking of the groups. Key variables represented in the dataset are `sex`, `age_group`, `eop_size_mm` and its category ranking `eop_size_category`, and `fhp_size_mm` and its category ranking `fhp_size_category`. There were `r nrow(exo_df)` participants included in the study, where `r nrow(filter(exo_df, sex == "Female"))` were women and `r nrow(filter(exo_df, sex == "Male"))` were men (Table 1). A review of the dataset revealed potential inconsistencies between categorical variables and their continuous counterparts. For the variable `eop_size_category`, the 4th categories' EOP sizes ranged from 2.11mm to 35.7mm, which is inconsistent with the intended ranges (Fig. 1).
  

```{r}
kable(sum_table, format = "markdown")
```
**Table 1**

```{r}
# checking issues btwn categorical and continuous versions

exo_df |> 
  group_by(age_group) |> 
  summarize(
    min_age = min(age, na.rm = TRUE),
    max_age = max(age, na.rm = TRUE)
  )

exo_df |> 
  group_by(eop_size_category) |> 
  summarize(
    mean_mm = mean(eop_size_mm, na.rm = TRUE),
    min_mm = min(eop_size_mm, na.rm = TRUE),
    max_mm = max(eop_size_mm, na.rm = TRUE)
  )

exo_df |> 
  group_by(fhp_category) |> 
  summarize(
    mean_fhp = mean(fhp_size_mm, na.rm = TRUE),
    min_fhp = min(fhp_size_mm, na.rm = TRUE),
    max_fhp = max(fhp_size_mm, na.rm = TRUE)
  )
```

```{r, include=FALSE}

# visualize the inconsistency with EOP category

inconsistency_plot = 
  exo_df |> 
  ggplot(aes(x = eop_size_category, y = eop_size_mm)) +
  geom_boxplot() +
  labs(title = "Inconsistencies in EOP Variables",
       x = "EOP Category", y = "EOP Size (mm)") + 
  theme_minimal()
```


```{r}
inconsistency_plot
```

**Figure 1**

```{r include = FALSE}
fhp_data <- exo_df |> 
  filter(!is.na(fhp_size_mm))
```

```{r include=FALSE}
# create binary variable for EEOP (2-5)
exo_df <- exo_df |> 
  mutate(enlarged_eop = ifelse(eop_size_category >= 2, 1, 0))

# proportion per age_group * sex
eop_summary <- exo_df |> 
  group_by(age_group, sex) |> 
  summarize(
    n_total = n(),
    n_enlarged = sum(enlarged_eop, na.rm = TRUE), 
    prop_enlarged = n_enlarged / n_total, 
    .groups = "drop"
  )
```


```{r}
# create fhp plot

fhp_dist <- fhp_data |>
  ggplot(aes(x = sex, y = fhp_size_mm, fill = sex)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  geom_boxplot(width = 0.12, outlier.shape = NA, fill = "white") +
  facet_wrap(~ age_group, ncol = 3) +
  scale_fill_viridis_d(option = "D", alpha = 0.9) +
  labs(
    title = "Distribution of FHP Size by Sex and Age Group",
    x = "Sex",
    y = "FHP Size (mm)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold", hjust = 0.5)
  )
  
```

```{r}
# create EEOP plot

eeop_dist <- eop_summary |> 
  ggplot(aes(
    x = age_group, y = prop_enlarged, fill = sex)) +
  geom_bar(
    stat = "identity", position = position_dodge()) + 
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1)) + 
  scale_fill_viridis_d(alpha = 1) +
  labs(
    title = "Proportion of Enlarged EOP by Age Group and Sex",
    x = "Age Group", y = "Proportion EEOP") + 
  theme_minimal() + 
  theme(legend.position = "none",
        plot.title = element_text(face = "bold", hjust = 0.5))

combined_plot <- fhp_dist / eeop_dist  +
  plot_layout(heights = c(2,1))
```


```{r}
# combined plot
combined_plot
```

**Figure 2**

```{r, warning = FALSE}

# faceted scatter plot

scatterplot <- exo_df |> 
  ggplot(aes(
    x = fhp_size_mm, 
    y = eop_size_mm, 
    color = sex)) + 
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  scale_color_viridis_d(option = "D") +
  facet_grid(sex ~ age_group) +

  labs(
    title = "Association between FHP Size and EOP Size",
    x = "FHP Size (mm)",
    y = "EOP Size (mm)"
  ) + 
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold", hjust = 0.5)
  )
```


```{r warning = FALSE}
scatterplot
```

**Figure 3**

Consistent with Shaher & Sayers, Figure 2 shows that EEOPs are most prevalent among younger males, with the greatest proportion observed in the 18-30 year category. Median FHP size was slightly larger in males, with greater overall variability than in females. Figure 3 indicates a generally positive association between FHP and EOP sizes across most age groups and sexes.

The author's sample sizes in each age group were not fully consistent with the data available. For instance, the paper states that the 18-30-year group included 300 participants, whereas this dataset contains `r nrow(filter(exo_df, age_group == "18-30"))` individuals in that category. The mean FHP in the male cases was `r exo_df |> filter(sex == "Male") |> summarize(mean_fhp = round(mean(fhp_size_mm, na.rm = TRUE), 1)) |> pull(mean_fhp)` ± `r exo_df |> filter(sex == "Male") |> summarize(sd_fhp = round(sd(fhp_size_mm, na.rm = TRUE), 1)) |> pull(sd_fhp)` mm, consistent with the author's findings. Similarly, the mean FHP among females was `r exo_df |> filter(sex == "Female") |> summarize(mean_fhp = round(mean(fhp_size_mm, na.rm = TRUE), 1)) |> pull(mean_fhp)` ± `r exo_df |> filter(sex == "Female") |> summarize(sd_fhp = round(sd(fhp_size_mm, na.rm = TRUE), 1)) |> pull(sd_fhp)` mm. EEOP is defined by Shaher and Sayers as an EOP exceeding 10 mm in size. Using the corresponding variable `eop_size_category`, EEOP was defined defined as `eop_size_category >= 2.` Using this criteria, the prevalence of EEOP was `r exo_df |> summarize(prop_enlarged = mean(eop_size_category >= 2, na.rm = TRUE) * 100) |> pull(prop_enlarged) |> round(1)`, closely aligning with the reported value of 33% in the original study.

```{r}
fhp_data |> 
  summarize(
    mean_fhp = mean(fhp_size_mm, na.rm = TRUE),
    sd_fhp   = sd(fhp_size_mm, na.rm = TRUE)
  )

fhp_data |> 
  group_by(sex) |> 
  summarize(
    mean_fhp = mean(fhp_size_mm, na.rm = TRUE),
    sd_fhp   = sd(fhp_size_mm, na.rm = TRUE),
    n        = n()
  )

```


```{r}
exo_df |> 
  summarize(
    prevalence_enlarged_eop = mean(eop_size_category >= 2, na.rm = TRUE)
  )

exo_df |>
  filter(age_group == "60+", !is.na(fhp_size_mm)) |>
  summarize(
    percent_over_40 = mean(fhp_size_mm > 40) * 100
  )


```

The percentage of participants aged 60+ with FHP greater than 40 mm was calculated as `r exo_df |> filter(age_group == "60+", !is.na(fhp_size_mm)) |> summarize(mean(fhp_size_mm > 40) * 100) |> pull() |> round(1)`%, slightly smaller than the report's 34.5%. My findings support the authors' observation that males tend to exhibit more robust cranial traits; however, there is insufficient evidence to conclude that cell phone use causes horn growth. Given the retrospective, cross-sectional nature of the data, establishing such a link would require longitudinal and behavioral measures of posture and device usage.